---
title: "Hoops Longwing Data Analysis"
author: "Zane Billings"
date: "15 November 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
if (require('tidyverse')==FALSE) {install.packages('tidyverse', repos="https://cran.rstudio.com")}
if (require('GGally')==FALSE) {install.packages('GGally', repos="https://cran.rstudio.com")}
```

In order to start analyzing the Hoops' Longwing sample data, we will first load the `tidyverse` package suite. After loading the packages we need, we can use `readr::read_csv()` to load in the data. But, notice the imported data frame has a useless column at the beginning, which we can easily remove manually.

```{r initialize}
library(tidyverse)

butterfly <- read_csv("hoops_longwing_study.csv")
butterfly <- butterfly[ , -1]
```

Now that we have the data imported, we can go ahead and take a quick look at the summary and structure.
```{r summary}
summary(butterfly)
str(butterfly)
```

The only real change we need to make is to convert the `population` variable into a factor, since the functions provided in `readr` do not coerce strings to factors by default.

```{r pop factor}
butterfly$population <- as.factor(butterfly$population)
summary(butterfly$population)
```

Now that all of our variables are imported correctly, let's go ahead and document what each variable means.

\begin{table}[h!]
\centering
\begin{tabular}{l | l}
Variable & Description \\
\hline
Wing length & \\
Wing width & \\
Age & \\
Number of offspring & \\
Feeding range & \\
Color peak & \\
Number of mates & \\
Avg. scale size & \\
Antenna length & \\
Number of spots & \\
Population & \\
Dispersal distance & \\
Body length & \\
Sample ID & 
\end{tabular}
\end{table}
