---
title: "Hoops Longwing Sample Data Visualization"
author: "Zane Billings"
date: "15 November 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
if (require('tidyverse')==FALSE) {install.packages('tidyverse', repos="https://cran.rstudio.com")}
if (require('GGally')==FALSE) {install.packages('GGally', repos="https://cran.rstudio.com")}
if (require('pander')==FALSE) {install.packages('Pander', repos="https://cran.rstudio.com")}
if (require('ggcorrplot')==FALSE) {install.packages('Pander', repos="https://cran.rstudio.com")}
```

In order to start analyzing the Hoops' Longwing sample data, we will first load the `tidyverse` package suite. After loading the packages we need, we can use `readr::read_csv()` to load in the data. But, notice the imported data frame has a useless column at the beginning, which we can easily remove manually.

```{r initialize}
library(tidyverse)

butterfly <- read_csv("hoops_longwing_study.csv")
butterfly <- butterfly[ , -1]
```

Now that we have the data imported, we can go ahead and take a quick look at the summary and structure.
```{r summary}
summary(butterfly)
```

\vfill\newpage

```{r str}
str(butterfly)
```

The only real change we need to make is to convert the `population` variable into a factor, since the functions provided in `readr` do not coerce strings to factors by default.

```{r pop factor}
butterfly$population <- as.factor(butterfly$population)
summary(butterfly$population)
```

\vfill\newpage

So, now we can start exploring our data. Let's start by making histograms of all the numeric data.

```{r histograms}
# Sturges Rule calculation for number of histogram bins.
num_bins <- ceiling(log2(nrow(butterfly))) + 1
butterfly %>% 
  select(-population) %>% 
  gather(key = "field", value = "value", -"sample_id") %>% 
  ggplot(aes(x = value)) +
  geom_histogram(col = "black", 
                 fill = "white", 
                 position = "identity",
                 bins = num_bins) +
  facet_wrap(~field, scales = "free")
```

\vfill\newpage

We can also visualize the distributions of the data using density curves, although we have to adjust the interpolation settings because part of our data is integers.

```{r densitycurves}
# An alternative way to view the data
butterfly %>% 
  select(-population) %>% 
  gather(key = "field", value = "value", -"sample_id") %>% 
  ggplot(aes(x = value)) +
  geom_density(adjust = 2) +
  facet_wrap(~field, scales = "free")
```

\vfill\newpage

Now we have one categorical variable, so let's look at all of our data stratified by the `population` value.
```{r violins}
butterfly %>%
  gather(key = "field", value = "value", -c(sample_id, population)) %>% 
  ggplot(aes(x = population, y = value, fill = population)) +
  geom_violin(adjust = 2) +
  facet_wrap(~field, scales = "free") + 
  theme(legend.position = "bottom")
```

\vfill\newpage

Now, using the `GGally` package, we can also make a scatterplot matrix like we did with `graphics::pairs()`. In order to get this to be visible, we'll need to split up the data.

```{r pairs1}
library(GGally)

butterfly[1:100, 1:6] %>%
  ggpairs(aes(alpha = 0.2))
```

\vfill\newpage

For the second half of the data, we also need to exclude our non-numeric variables.

```{r pairs2}
butterfly[1:100, 7:16] %>% 
  select(-c(sample_id, population, latitude, longitude)) %>%
  ggpairs(aes(alpha = 0.2))
```

\vfill\newpage

As you can probably see, this visualization is not ideal when we have a lot of data.

Let's try a correlation table as well. However, note that while a correlation table can give us a good sense of linaer relationships, we lose any information we had about nonlinear relationships, which we have to examine visually if we don't have a hypothesis about their existence.

```{r corrtable}
library(pander)
butterfly %>% 
  select(-c(sample_id, population)) %>% 
  cor() %>% 
  pander()
```

\vfill\newpage

Or, we can use a function from the `ggcorrplot` library to make a visual representation of the same data.

```{r corrplot}
library(ggcorrplot)
butterfly %>% 
  select(-c(sample_id, population)) %>%
  cor() %>% 
  ggcorrplot()
```

\vfill\newpage

Just an example of a scatterplot.

```{r scatterplot}
butterfly %>% 
  select(wing_length, wing_width) %>% 
  ggplot(aes(x = wing_length, y = wing_width)) +
  geom_point(alpha = 0.2) + 
  geom_smooth(method = "lm", lty = 2) +
  labs(x = "Wing length (cm)", y = "Wing width (cm)") +
  theme_minimal()
```

\vfill\newpage

We can also do a LOT with scatterplots.

```{r adv_scatter1}
butterfly %>% 
  select(wing_length, wing_width, population) %>% 
  ggplot(aes(x = wing_length, 
             y = wing_width, 
             col = population)) +
  geom_point(alpha = 0.2) + 
  geom_smooth(method = "lm")
```

\vfill\newpage

There's literally so many aesthetic arguments you can use for a scatterplot (I contend that not all of them are useful).

```{r adv_scatter2}
butterfly %>% 
  select(wing_length, wing_width, population, antenna_length, body_length) %>%
  ggplot(aes(x = wing_length,
             y = wing_width,
             col = antenna_length,
             size = body_length)) +
  geom_point(alpha = 0.2) + 
  facet_wrap(~population)
```